---
title: "Predicting Income - HarvardX Capstone Project"
author: "Kevin Jané"
date: "2022"
output:
  bookdown::pdf_document2:
    keep_tex: true
    number_sections: true
    toc: true
    toc_depth: 3
    latex_engine: lualatex
  rmarkdown::pdf_document:
     keep_tex: true
     number_sections: true
     toc: true
     toc_depth: 3
     latex_engine: xelatex
documentclass: report
fontsize: 12pt
links-as-notes: true
linkcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache=TRUE, cache.lazy = FALSE, warning = FALSE)
```

\newpage
# Preface - Introduction

The purpose of this capstone project is to create a predictive project to achieve the
[Professional Certificate of the Data Science](https://www.edx.org/es/professional-certificate/harvardx-data-science) 
courses taught by Harvard University.

The data science job market is exponentially growing being in the 
[top 3 of jobs most sought after](https://www.forbes.com/sites/forbeshumanresourcescouncil/2021/05/20/hr-leaders-share-14-in-demand-skills-employers-want-in-2021/?sh=44ba748d1e45), 
this can allow us to infer that the world is giving so much importance to open data than 
it was years ago, recognizing the potential of data analysis and prediction models for 
the global social-economic development.

I as an undergraduate economics student, being passionate about data, being able to manipulate 
data with R facilitates doing data analyses. 
Also, predictive models are essential to our, which become more time-efficient with R.

My lovely small country, Paraguay, in the center of South America, also called the "heart of South America".
A country that has been growing these last years, but the needing of data, people who analyse the data and who
investigate the behavior of the economy and everything that is going on in the country, created the my love for 
the data analysis.


In this project well going to use the [1994 Census Income Data Set](http://www.census.gov/ftp/pub/DES/www/welcome.html), 
that is a dataset donated by Ronny Kohavi and Barry Becker and provided by the UCI Machine Learning Repository.
One variable is related with income, our goal in this project is trying to predict the income based on data from the Census database.

\newpage
# Exploratory Data Analysis
## Data Preparation
In this section, we install and load every packages required for this project, as well as the 1994 Census database provided.

```{r, warning=FALSE, message=FALSE}
if(!require(randomForest)) install.packages("randomForest")
if(!require(reldist)) install.packages("reldist")
if(!require(readxl)) install.packages("readxl")
if(!require(dplyr)) install.packages("dplyr")
if(!require(tidyr)) install.packages("tidyr")
if(!require(dslabs)) install.packages("dslabs")
if(!require(stringr)) install.packages("stringr")
if(!require(forcats)) install.packages("forcats")
if(!require(ggplot2)) install.packages("ggplot2")
if(!require(caTools)) install.packages("caTools")
if(!require(rpart.plot)) install.packages("rpart.plot")
if(!require(ISLR)) install.packages("ISLR")
if(!require(e1071)) install.packages("e1071")
if(!require(OneR)) install.packages("OneR")
if(!require(tidyverse)) install.packages("tidyverse", 
                                         repos = "http://cran.us.r-project.org")
if(!require(ggthemes)) install.packages("ggthemes", 
                                        repos="http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", 
                                     repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", 
                                          repos = "http://cran.us.r-project.org")
if(!require(rpart)) install.packages("rpart", 
                                     repos="http://cran.us.r-project.org")
if(!require(MLmetrics)) install.packages("MLmetrics", 
                                         repos="http://cran.us.r-project.org")
if(!require(haven)) install.packages("haven", 
                                     repos="http://cran.us.r-project.org")
```

```{r, warning=FALSE, message=FALSE}
library(randomForest)
library(reldist)
library(readxl)
library(dplyr)
library(tidyr)
library(dslabs)
library(stringr)
library(forcats)
library(ggplot2)
library(tidyverse)
library(OneR)
library(caret)
library(data.table)
library(ggthemes)
library(caTools)
library(rpart)
library(ISLR)
library(e1071)
library(MLmetrics)
library(haven)
library(hrbrthemes)
library(viridis)
library(rpart.plot)
```

```{r, warning=FALSE, message=FALSE}
set.seed(1, sample.kind = "Rounding")
trainFileName = "adult.data"; testFileName = "adult.test"

if (!file.exists (trainFileName))
    download.file (
      url = "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.data", 
      destfile = trainFileName)

if (!file.exists (testFileName))
    download.file (
      url = "http://archive.ics.uci.edu/ml/machine-learning-databases/adult/adult.test",
      destfile = testFileName)
colNames = c ("age", "workclass", "fnlwgt", "education", 
              "educationnum", "maritalstatus", "occupation",
              "relationship", "race", "sex", "capitalgain",
              "capitalloss", "hoursperweek", "nativecountry",
              "incomelevel")

adult = read.table (trainFileName, header = FALSE, sep = ",",
                       strip.white = TRUE, col.names = colNames,
                        na.strings = "?", stringsAsFactors = TRUE)
```

```{r, warning=FALSE}
str(adult)
```

The adult dataset contains **32561** rows and **15** variables, wich are:
\begin{itemize}
  \item age <int>: continuous.
  \item workclass <Factor>: Private, Self-emp-not-inc, Self-emp-inc, Federal-gov, Local-gov, State-gov, Without-pay, Never-worked.
  \item fnlwgt <int>: continuous.
  \item education <Factor>: Bachelors, Some-college, 11th, HS-grad, Prof-school, Assoc-acdm, Assoc-voc, 9th, 7th-8th, 12th, Masters, 1st-4th, 10th, Doctorate, 5th-6th, Preschool.
  \item educationnum <int>: continuous.
  \item maritalstatus <Factor>: Married-civ-spouse, Divorced, Never-married, Separated, Widowed, Married-spouse-absent, Married-AF-spouse.
  \item occupation <Factor>: Tech-support, Craft-repair, Other-service, Sales, Exec-managerial, Prof-specialty, Handlers-cleaners, Machine-op-inspct, Adm-clerical, Farming-fishing, Transport-moving, Priv-house-serv, Protective-serv, Armed-Forces.
  \item relationship <Factor>: Wife, Own-child, Husband, Not-in-family, Other-relative, Unmarried.
  \item race <Factor>: White, Asian-Pac-Islander, Amer-Indian-Eskimo, Other, Black.
  \item sex <Factor>: Female, Male.
  \item capitalgain <int>: continuous.
  \item capitalloss <int>: continuous.
  \item hoursperweek <int>: continuous.
  \item nativecountry <Factor>: United-States, Cambodia, England, Puerto-Rico, Canada, Germany, Outlying-US(Guam-USVI-etc), India, Japan, Greece, South, China, Cuba, Iran, Honduras, Philippines, Italy, Poland, Jamaica, Vietnam, Mexico, Portugal, Ireland, France, Dominican-Republic, Laos, Ecuador, Taiwan, Haiti, Columbia, Hungary, Guatemala, Nicaragua, Scotland, Thailand, Yugoslavia, El-Salvador, Trinadad Tobago, Peru, Hong, Holand-Netherlands.
  \item incomelevel <Factor>: >50K, <=50K.
\end{itemize}

**NA values**

```{r, warning=FALSE}
na_v <- sapply(adult, function(x) sum(is.na(x)))
na_v
```

**Percentage of NA values**

```{r}
pna_v <- sapply(adult, function(adult){sum(is.na(adult)==T)*100/length(adult)})
round(pna_v, digits = 3)
```

We can see that the variables ```workclass``` (**5.639%**),```occupation```(**5.660%**) and ```nativecountry```(**1.790%**) have NAs.
Actually, this is not a good thing because these variables could be a very good predictors of income.

So, we want to remove all the NAs from the dataset

```{r, warning=FALSE}
adult = adult[!is.na(adult$workclass) & !is.na(adult$occupation),]
adult = adult[!is.na(adult$nativecountry),]
adult$fnlwgt = NULL
```

## Data Analysis

The variables/columns used in ```adult``` are:

\begin{itemize}
  \item age <int>
  \item workclass <Factor>
  \item education <Factor>
  \item educationnum <int>
  \item maritalstatus <Factor>
  \item occupation <Factor>
  \item relationship <Factor>
  \item race <Factor>
  \item sex <Factor>
  \item capitalgain <int>
  \item capitalloss <int>
  \item hoursperweek <int>
  \item nativecountry <Factor>
\end{itemize}

We want to see the distribution of income between the variables, we can plot it and see their behavior.

In the next plot we see the frequency of ages in the database, with the condition of the income.

```{r, warning=FALSE}
adult %>%
  ggplot(aes(age)) +
  geom_histogram(aes(fill=incomelevel), binwidth = 0.5) +
  theme_bw() + xlab("Age") + ylab("Frecuency") +
  ggtitle("Frecuency of ages for income levels")
```

In the next plot, we graph the frequency of years of education with the condition of the level of income.

We actually see that since 9 years of studying dedication, that is a high school degree, 
the frequency of people who earns **<=50K** are the predominant. But also there is more frequency
of people who earns **>50K** than 8 years of study or earlier.

```{r, warning=FALSE}
adult %>%
  ggplot(aes(educationnum)) +
  geom_histogram(aes(fill=incomelevel), binwidth = 0.5) +
  scale_y_log10() + theme_bw() + 
  xlab("Years of Education") + ylab("Frequency") +
  ggtitle("Frecuency of years of education for income levels")
```

We can see that there is more frequency of people who works in a 40 hour job. And we can see that at every level, is more frequently to find people
who earns less than 50k (**<=50K**)

```{r, warning=FALSE}
adult %>%
  ggplot(aes(hoursperweek)) +
  geom_histogram(aes(fill=incomelevel), binwidth = 2.5) + 
  scale_y_log10() + theme_bw() + 
  xlab("Hours per Week") + ylab("Frequency") +
  ggtitle("Frecuency of hours per week for income levels")
```

If we want to see how much do people earn based on the country that they are from, we see that as the last plot, the behavior is very similar.
At very level or country, we see that it is more common to see more people that earns less than 50k. (**<=50K**), but in the case of United States,
we see that there is more people who earns more than 50k (**>50k**)

```{r}
adult %>%
  ggplot(aes(x=reorder(nativecountry, nativecountry, function(x) length(x)))) +
  geom_bar(aes(fill=incomelevel), width = 0.8, position = "identity") +
  scale_y_log10() + theme_bw() + 
  xlab("Native Country") + ylab("Frequency") +
  ggtitle("Frecuency of countries for income levels") +
  coord_flip()
```

## Variables for modeling

After the initial data exploration, we want to select at least three variables for the income prediction.

So, for the predictions, we are going to use the next variables:

\begin{itemize}
  \item age <int>
  \item education <Factor>
  \item occupation <Factor>
  \item race <Factor>
  \item sex <Factor>
\end{itemize}

## Methology for modeling

We are going to use three models in this project, those are:
\begin{itemize}
  \item SVM (Support Vector Machine)
  \item Decision Tree
  \item Random Forest
\end{itemize}

For evaluating those models, we are going to use four metrics, those are:
\begin{itemize}
  \item Accuracy
\[
\frac{\text{true positives} + \text{true negatives}}{\text{true positives} + \text{false negatives} + \text{true negatives} + \text{true negatives}}
\]
  \item Sensitivity
\[
\frac{\text{true positives}}{\text{true positives} + \text{false negatives}}
\]
  \item Specificity
\[
\frac{\text{true negatives}}{\text{true negatives} + \text{false positives}}
\]
  \item F1 Score
\[
2 \times \frac{\text{precision} \times \text{recall}}{\text{precision} + \text{recall}}
\]
\end{itemize}

The summary of the results with all the metrics are going to be in the results section.

\newpage
# Modeling

For this project, we have to divide ```adult``` database into ```train_set``` and ```test_set```. ```train_set``` 
is used to create all the models and ```test_set``` is used to prove how nice those models works.

```{r}
# Sample, train and test sets for the models
sample.adult <- sample.split(adult$incomelevel, SplitRatio = 0.80)
train_set = subset(adult, sample.adult == TRUE)
test_set = subset(adult, sample.adult == FALSE)
```

## SVM (Support Vector Machine)

This is a supervised model known as Support Vector Machine.

This is a classification algorithm, with the objective of finding a 
hyperplane that separates data points of one class from those of another class. Basically, this works on the principle of
a maximum marginal classifier.
\newline Source: [Math Works](https://www.mathworks.com/discovery/support-vector-machine.html)

```{r}
# Applying SVM Model
svm.adult = svm(incomelevel ~
                  age+education+occupation+race+sex,
                data = train_set)
# Prediction of data and Confusion Matrix
test_set$pred.value = predict(svm.adult, newdata=test_set, type="response")
model1 <- table(test_set$income, test_set$pred.value)
confusionMatrix(model1)
```

```{r}
F1_Score(test_set$income, test_set$pred.value)
```

We add the results of this model to a data frame.

```{r}
results<- data.frame(
  Model="SVM (Support Vector Machine)",
  Accuracy=
    Accuracy(test_set$income, 
             test_set$pred.value),
  F1Score=
    F1_Score(test_set$income,
             test_set$pred.value),
  Sensitivity=
    sensitivity(test_set$income,
                test_set$pred.value),
  Specificity=
    specificity(test_set$income,
                test_set$pred.value))
results
```

We can see that with this model we have a really good accuracy and a f1 score, but a little low specificity.

## Decision Tree

This model that we are going to apply in this case is a one step decision tree.
This model is harder to interpret but has an accuracy a little better than the linear regression. It goes thru
the different variables to see which bracked it ends.
\newline Source: [Cran Project](https://cran.r-project.org/web/packages/OneR/index.html)


```{r}
# Applying Decision Tree Model
detree <- rpart(incomelevel ~
                  age+education+occupation+race+sex,
                data = train_set)
# Prediction of data and Confusion Matrix
test_set$pred.value2 = predict(detree, newdata=test_set, type="class")
model2 <- table(test_set$income, test_set$pred.value2)
confusionMatrix(model2)
```

```{r}
F1_Score(test_set$income, test_set$pred.value2)
```

```{r}
results<- bind_rows(
  results,
  data.frame(Model="Decision Tree",
             Accuracy=Accuracy(test_set$income,
                               test_set$pred.value2),
             F1Score=F1_Score(test_set$income,
                              test_set$pred.value2),
             Sensitivity=sensitivity(test_set$income,
                                     test_set$pred.value2),
             Specificity =specificity(test_set$income,
                                      test_set$pred.value2)))
results
```

In this case, we see that our specificity improved, we can try another model to see how it behave.

## Random Forest

This model consist of a large number of individual decision trees that operate as an ensemble.
Each individual tree in the random forest splits out a class prediction and the class with the most votes becomes our model's prediction.
\newline Source: [Towards Data Science](https://towardsdatascience.com/understanding-random-forest-58381e0602d2)

```{r}
set.seed(4543) # this is for reproducibility
# Applying Random Forest Model
rfmodel <- randomForest(incomelevel ~ 
                          age+education+occupation+race+sex,
                        data = train_set, importance = TRUE)
# Prediction of data and Confusion Matrix
test_set$pred.value3 = predict(rfmodel, newdata=test_set)
model3 <- table(test_set$income, test_set$pred.value3)
confusionMatrix(model3)
```

```{r}
F1_Score(test_set$income, test_set$pred.value3)
```

```{r}
results<- bind_rows(
  results,
  data.frame(Model="Random Forest",
             Accuracy=Accuracy(test_set$income,
                               test_set$pred.value3),
             F1Score=F1_Score(test_set$income,
                              test_set$pred.value3),
             Sensitivity=sensitivity(test_set$income,
                                     test_set$pred.value3),
             Specificity =specificity(test_set$income,
                                      test_set$pred.value3)))
results
```

For this final model, we see that the specificity is a little lower too, but a really nice accuracy and f1 Score.

\newpage
# Results

This is a summary of the results of all the models that we did before.
\newline All of these models were trained on ```train_set``` (**80%** of adult database) and validated with ```test_set``` (**20%** of adult database).

```{r}
results
```

**Decision Tree** is the best model if we look at the ```F1 Score``` and the ```specificity```.
\newline But, **Random Forest** is the best if we look at the ```accuracy``` and the ```sensitivity```.
\newline In this case **SVM (Support Vector Machine)** had the lowest percentages in the indicators, being the worst among them.


# Conclusion

As a first step, we loaded de "Adult" or "Census+Income" database from the 1994.
We split it into two parts, one for training (80%) and the other one for testing (20%).

After the exploration we proceed to model the algorithms. 

**Limitations**

We actually used only three types of models, and this project can be used for a more rigorous machine learning project.

**Future Work**

As mentioned, this project can be used for a more rigorous machine learning project.
\newline Other thing that can be done in the future is the database, this is from the 1994, 
this kind of investigations can be very useful for reports and education sources for
machine learning and data analysis for timelines analysis and predictions.
\newline As well, the Specificity and the Sensitivity can be prioritize one or another based on
the type of policy we want to make. And, considering the consequences of making one type of error or
another, we'll know which type of error is more severe or costly than making the other type of error.
We can make clear and choose what type of error(type 1 or 2) based on which one
have more significance and power for the test.






